On The Subspace Learning for Network Attack Detection

### 2019-06-14 docs/thesis/chapters/4_m_rpca/m_rpca.tex
after fix bold notation

Some widely adopted algorithms for anomaly detection assume a gaussian distributed data, however this assumption may not be observed in some real world problems, such as the case of network traffic analysis, where network traffic features are usually more characterized by skewed and heavy-tailed distributions \cite{lakhina2005mining,benson2010network}.

@@

This findings show that the skeweness and heavy-tailed distributions may be important for network traffic analysis, and can motivate evaluations to study the impact of skewed data into algorithms that rely on gaussian distributed data, as well as it can highlight opportunities to  exploit the skeweness and heavy-tailed distributions in order to obtain accurate classifiers for network anomaly detection.

@@

We believe that the skewness of anomalous and normal traffic can highlight features for improving anomaly detection in imbalanced data, 
and that the distance between robust estimates of normal traffic and anomalies can highlight discrepancies and be used for network attack detection. Therefore, we propose the md-RPCA, which is an approach based on distances from moments computed from robust subspace learned by Robust Principal Component Analysis, in order to detect anomalies from network traffic. The proposed approach relies on a robust subspace of normal traffic for estimating higher moments. The anomaly detection from new observations evaluate the Mahalanobis distance between the robust higher moments and the higher moments of new observations, in a semi-supervised or unsupervised fashion, without the computational cost of new robust subspace learning for new observations.

We evaluate the accuracy of the md-RPCA for anomaly detection on synthetic data set and on the CTU-13 data set \cite{garcia2014empirical}, which is a large data set of normal, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}. The Experimental evaluation compares our proposal to widely adopted algorithms for anomaly detection based on clustering and statistical approaches.

% We also evaluate the results of ROBPCA \cite{hubert2005robpca}, which is a method that also relies on robust estimates with adjusted outlyingness based on robust skewness.

The main contribution of this work is a novel semi-supervised and unsupervised method for anomaly detection in skewed and imbalanced data, with results of experimental evaluation on synthetic and real data sets.

This Chapter is organized as follows. In Section \ref{sec:4_relatedworks}, it is conducted a literature review of network anomaly detection, botnet detection, and imbalanced learning. Section \ref{sec:4_datamodel} presents the data model and the evaluated data set. Section \ref{sec:4_proposal} describes the proposed approach for network attack detection. Section \ref{sec:4_experiments} discusses the experimental validation and Section \ref{sec:4_results} presents the results. Finally, Section \ref{sec:4_conclusion} draws the conclusions and the suggestions for future work.

@@

% TODO - re-write this, it is strange
% network anomaly detectiom
Bhuyan \emph{et al.} \cite{bhuyan2014network} provide an overview of facets of network anomaly detection, present attacks normally encountered by network intrusion detection systems, and categorize existing network anomaly detection methods and systems based on the underlying techniques. Ahmed \emph{et al.} \cite{ahmed2016survey} present an analysis of four major categories of anomaly detection techniques which include classification, statistical, information theory and clustering. Moustafa \emph{et al.} \cite{moustafa2019holistic} discuss aspects of anomaly-based Network Intrusion Detection Systems (NIDSs), explains cyber-attacks and new solutions for anomaly detection, and provides a benchmark data sets for training and validating approaches for detection.

@@

Acarali \emph{et al.} \cite{acarali2016survey} surveys network-based detection approaches for HTTP-based botnets, surveys traffic-based features used to detect bot traffic and presents an abstraction of the main types of features related to protocols and OSI layers. Wang \emph{et al.} \cite{Wang2018ddosbotnetssurvey} present an analysis based on 50,704 different Internet DDoS attacks originated of 674 botnets from 23 different botnet families with a total of 9,026 victim belonging to 1,074 organizations in 186 countries. Their analysis reveals that geolocation of the attacking sources follows patterns and enables source prediction, and highlights that multiple attacks to the same target also exhibit strong patterns of inter-attack time interval, also presents that there is a trend for different botnets to launch DDoS attacks targeting the same victim, simultaneously or in turn.

@@

% justification for the data set and present who else use it
Several approaches for network attack detection uses the KDD 99 \cite{ahmed2016survey,osanaiye2016distributed,bhuyan2014network} data sets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 data set are criticized by the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled data sets currently in use today by researchers \cite{osanaiye2016distributed,bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} data set is the refined version of the KDD 99 data set that redundant data records are removed, in order to avoid biased classifications. However, NSL-KDD data set maintain the limitations of the KDD 99 regarding volume and reproduction of network traffic similar to the real networks.

The objective of Garcia \emph{et al.} \cite{garcia2014empirical} is to compare three botnet detection methods using a simple and reproducible methodology, a good data set and a new error metric. This paper evaluates some data sets for network anomaly detection, surveys some approaches for botnet detection, proposes two methods (BClus and CAMNEP) for botnet detection and compare results to BotHunter \cite{gu2007bothunter}.

Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by attack, normal and background labeled data, in an imbalanced distribution like in a real network. The authors recommend scenarios for training and testing, in the way that none of the botnet families used in the training and cross-validation data set should be used in the testing data set, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Adopting the training and testing approach proposed by Garcia \emph{et al.}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training.

Wang and Paschalidis \cite{wang2017botnet} proposed a botnet detection approach based on anomaly and community detection, aiming for detecting botnets and identifying bots before the botnet becomes active. The first stage detects anomalies by leveraging large deviations of an empirical distribution. The second stage detects the bots using ideas from social network community detection in a graph that captures correlations of interactions among nodes over time. This work is compared with the BotHunter \cite{gu2007bothunter} on the CTU-13 botnet data set \cite{garcia2014empirical}.

% \cite{da2018online} proposes an approach using the Very Fast Decision Tree, a classification algorithm used on stream mining that can learn incrementally when needed, to identify botnets by observing network flows. When evaluating the approach on multiple scenarios with different botnets, we were able to achieve high performance metrics on the majority of scenarios, while using a significantly low number of labelled instances. It discard scenarios 5 and 6. They does not compute F1, but it is easy to compute it from precision and recall provided.

% Anomaly detection by subspace larning
Traditional PCA-based anomaly detection models are not suitable for anomaly interpretation, as they judge whether a data instance is an anomaly or not based on the length of its projection on the abnormal subspace spanned by the less significant principal components, and there is no direct mapping between PCA’s dimensionality-reduced subspace and the original feature space \cite{ringberg2007sensitivity}. However, to overcome the above mentioned limitations, some approaches based on PCA have been proposed for network anomaly detection. Callegari \emph{et al.} \cite{callegari2011novel} proposed a PCA-based method for identifying the network traffic flows responsible for an anomaly detected at the aggregate level, by means of a separation of normal and anomalies according to principal components (normal) and remaining (anomalies). Lee \emph{et al.} \cite{Lee2013} presented OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. Vieira \emph{et al.} \cite{vieira2017model} proposed a framework that applies Model Order Selection (MOS) for detection of time frames under attack and uses similarity analysis to extract details and detect the time and ports under attack.

% Anomaly detection by robust subspace learning
The problem of PCA or subspace learning for outlier corrupted data is called Robust Principal Component Analysis (RPCA) or robust subspace learning \cite{candes2011robust, vaswani2018robust}. RPCA aims to be resilient to outliers by means of a robust subspace learning \cite{vaswani2018robust} for outlier corrupted data, decomposing a given data matrix $\textbf{X}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, which refers to outliers’ matrix. This definition is also referred to as the sparse + low rank (S + LR) formulation \cite{vaswani2018robust}. We refer to \cite{lerman2018overview} and \cite{vaswani2018robust} for more details regarding robust subspace learning.

RPCA is been mainly applied to computer vision, in problems of robust subspace tracking and robust subspace recovery. However, RPCA has also been adopted for general outlier detection \cite{hubert2005robpca,hubert2009robustskewed,cherapanamjeri2017thresholding,zhou2017anomaly,NetflixSurus} and for anomaly detection on network traffic \cite{pascoal2012robust}. ROBPCA \cite{hubert2005robpca} intends to identify outliers using PCA from robust estimates of mean and covariance matrix, to reduce the data dimensions and plotting the orthogonal distances versus the robust score distances, to flag an outlier map. However, ROBPCA flags many points as outlying when the original data is skewed. Therefore, Hubert \emph{et al.} \cite{hubert2009robustskewed} proposed ROBPCA-AO, which improves ROBPCA for problems with skewed data, by means of an adjusted outlyingness based on robust skewness.

@@

Mahalanobis Distance (MD) is a generalized distance which is useful for determining the similarity between an unknown sample and a collection of known samples, by considering the covariances between the variables and their mean values. The MD is a measure of the distance between a vector $\boldsymbol{x}$ and a distribution $\boldsymbol{X}$, introduced by P. C. Mahalanobis in 1936 \cite{mahalanobis1936md}. It is a multi-dimensional generalization for measuring how many standard deviations away $\boldsymbol{x}$ is from the mean $\boldsymbol{\bar{x}}$ and covariance $\boldsymbol{\bar{\Sigma}}$ of $\boldsymbol{X}$. MD has been used for distance based anomaly detection with robust estimates in many areas, assuming that the Mahalanobis Distance between robust estimates and new observations can reveal anomalies.

We propose an approach based on higher moments distances from robust subspace for botnet detection. The proposed approach relies on a robust subspace of normal traffic for estimating higher moments. Thus, we propose that the anomaly detection for new observations should evaluate the Mahalanobis distance between the robust higher moments and the higher moments of new observations, in a semi-supervised fashion and without the computational cost of new robust subspace learning for new observations, or in a unsupervised approach.

@@

In this paper, scalars are denoted by italic letters (\emph{a, b, A, B, $\alpha$, $\beta$}), vectors by lowercase bold letters (\textbf{a, b}), matrices by uppercase bold letters (\textbf{A, B}), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix \textbf{A}. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively. The Frobenius norm is denoted as $\left\| \mathord{\cdot} \right\|_F$, while $\left\| \mathord{\cdot} \right\|_*$ denotes the nuclear norm of a matrix and $\left\| \mathord{\cdot} \right\|_1$ means the sum of the absolute values of matrix entries. We also define the operator $\langle \mathord{\cdot} \rangle$, that is the standard trace inner product, and $[ \mathord{\cdot} ]^s$, which denotes the $n$ largest values of a vector.

This section also presents a description of the synthetic data model in Subsection \ref{sec:4_synthetic} and in Subsection \ref{sec:4_CTU-13} we describe the data model for the CTU-13 data set.

\subsection{The Synthetic data set}
\label{sec:4_synthetic}

In order to analyse the null hypotheses $H_{1rob.sub}$ and $H_{0rob.sub}$, which evaluate if the robust subspace learning can improves the anomaly detection from imbalanced and skewed data, we create two synthetic data sets characterized by skewed and heavy tailed distributions. We selected Pareto and Lognormal distributions to evaluate the anomaly detection capacity of our proposal, considering that these distributions are skewed and heavy-tailed, and have been adopted to characterize network traffic of the Internet and data centers \cite{benson2010network,leon2017probability}. We also evaluate our proposal and the selected algorithms for anomaly detection on gaussian distributed data set, in order to analyse the detection rate for not skewed and not heavy tailed distributions.

The addictive anomalies for Pareto and Lognormal distributions are gaussian distributed, while an uniform noise is used to add anomalies into the gaussian distributed data set. We generate $\boldsymbol{Y}_p$, $\boldsymbol{Y}_l$ and $\boldsymbol{Y}_g$ for Pareto, Lognormal and Gaussian multivariate distributions, respectively, with 2 features. Each contaminated data set is composed by a number of legitimate observations and the contaminated samples, defined by the adopted $c$.

The Pareto distribution contaminated by gaussian noise is denoted as $\boldsymbol{Y}_p^c$, while $\boldsymbol{Y}_l^c$ denotes the Lognormal distribution contaminated by gaussian noise, and $\boldsymbol{Y}_g^c$ denotes the contaminated gaussian data contaminated by an uniform distribution.

This synthetic data set simulate a total of 2400 events for each scenario, where the number of legitimate observations is simulated according to the desired contamination rate. We evaluate contamination rates $c$ between 1\% and 50\%, to simulate the imbalanced data of anomaly detection problems during our experiments.

The Figure \ref{fig:4.01} depicts the distribution of Pareto and gaussian contamination of one variable, as a visual example of $\boldsymbol{Y}_p^c$, while the Figure \ref{fig:4.02} shows one example of Lognormal contaminated by gaussian noise, generated for $\boldsymbol{Y}_l^c$.

\begin{figure}[!htb]
	\centering
	\subfigure[Pareto and Gaussian Anomalies]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/4_Xpg2.png}
		\label{fig:4.01}
	}
	\subfigure[Lognormal and Gaussian Anomalies]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/4_Xlogng2.png}
		\label{fig:4.02}
	}
	\caption[Examples of skewed and heavy tailed distributions]{Examples of skewed and heavy tailed distributions}
	\label{fig:4.03}
\end{figure}

@@

\begin{itemize}
	\item \textbf{Attacks:} Click Fraud (CF), Port Scan (PS), Compiled and Controlled by Authors (US), SPAM and DDOS;
	\item \textbf{C\&C:} IRC, P2P and HTTP.
\end{itemize}

@@

We conducted and Exploratory Data Analysis (EDA) on the CTU-13 and observed that some features are skewed and present an overlapping between normal and anomalous flows, as can be seen in Figure \ref{fig:4.05} and Figure \ref{fig:4.05}, that present the distributions of TCP state of scenario 16 and the type of services from destination of scenario 10.

\begin{figure}[!htb]
	\centering
	\subfigure[State of scenario 16.]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/raw_distplot_capture20110816_State.png}
		\label{fig:4.04}
	}
	\subfigure[Destination Type of service of scenario 10.]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/raw_distplot_capture20110810_dTos.png}
		\label{fig:4.05}
	}
	\caption[Skewness and Overlapping]{Example of skewness and overlapping}
	\label{fig:4.06}
\end{figure}

However, it is not possible to observe a pattern on distributions of all the features, as depicted by Figures \ref{fig:4.07} and \ref{fig:4.08}, that show the distributions of TCP states of the scenario 10 and source ports of scenario 16, and highlight the distributions of normal and anomalous flows.

\begin{figure}[!htb]
	\centering
	\subfigure[State of scenario 10.]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/raw_distplot_capture20110810_State.png}
		\label{fig:4.07}
	}
	\subfigure[Source Port of scenario 16.]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/raw_distplot_capture20110816_Sport.png}
		\label{fig:4.08}
	}
	\caption[Skewness and Overlapping]{Skewness and Overlapping}
	\label{fig:4.09}
\end{figure}

Due to the number of available features and the class overlapping between normal and anomalous flows, we performed an correlation analysis and cross validation in order to identify the best set of features for network attack detection. Therefore, we adopt the following features: state, destination type of service, destination port, source port, total number of packets, total number of bytes and number of bytes from the source.

\section{Moment Distances from Robust Subspace for Botnet Detection}
\label{sec:4_proposal}

This section describes the proposed approach for network attack detection by means of a distance analysis between moments computed from a robust subspace and new observations of network traffic.

Robust subspace learning can be defined as the decomposition of a given data matrix $\textbf{X}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, with outliers or noise. We propose to learn a robust subspace $\textbf{L}$ from the normal traffic $\textbf{X}^s$ for computing $\textbf{L}^s$, $\textbf{S}^s$ and the robust moments, i.e. the mean $\boldsymbol{\bar{x}}$, skewness $\boldsymbol{\bar{s}}$ and kurtosis $\boldsymbol{\bar{k}}$, in order to evaluate the distance $\boldsymbol{d}$ of new observations $\boldsymbol{X}^c$ to these moments and classify as anomalous the observations with the largest distances.

Robust PCA is a well known method to recover a low-rank matrix $\textbf{L}$ and sparse matrix $\textbf{S}$ from corrupted measurements modeled as $\textbf{X} = \textbf{L} + \textbf{S}$. This decomposition in low-rank and sparse matrices can be achieved by techniques such as Principal Component Pursuit method (PCP), and by optimization methods, such as the Augmented Lagrange Multiplier Method (ALM), Alternating Direction Method (ADM), Fast Alternating Minimization (FAM) or Iteratively Reweighted Least Squares (IRLS) \cite{candes2011robust,vaswani2018robust,lerman2018overview}.

>>>>

Some widely adopted algorithms for anomaly detection assume a Gaussian distributed data, however this assumption may not be observed in some real world problems, such as the case of network traffic analysis, where network traffic features are usually more characterized by skewed and heavy-tailed distributions \cite{lakhina2005mining,benson2010network, leon2017probability}.

@@

These findings show that the skeweness and heavy-tailed distributions may be important for network traffic analysis, and can motivate researches to evaluate the impact of skewed data into algorithms that rely on Gaussian distributed data. These findings can also highlight opportunities to  exploit the skeweness and heavy-tailed distributions to obtain accurate classifiers for network anomaly detection.

@@

We believe that the skewness of anomalous and normal traffic can highlight features for improving anomaly detection in imbalanced data. We also believe that the distance between robust estimates of normal traffic and contaminated observations can highlight discrepancies and be used for network attack detection.

Therefore, we propose the m-RPCA, which is an approach based on distances of moments computed from robust subspace learned by Robust Principal Component Analysis (RPCA), in order to detect anomalies from skewed data and network traffic. The proposed approach relies on a robust subspace computed from supposed normal observations, for estimating the moments to be used for distance analysis. The anomaly detection from new observations evaluate the Mahalanobis distance between the robust moments and the moments of new observations, in a semi-supervised fashion, without the computational cost of new robust subspace learning for new observations. The m-RPCA can also be computed as an unsupervised algorithm, with subspace learning from the same contaminated data that is the target of the anomaly detection analysis.

We evaluate the accuracy of the m-RPCA for anomaly detection on simulated data sets, with skewed and heavy-tailed distributions, and for the CTU-13 data set \cite{garcia2014empirical}, which is a large data set of normal, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}. The Experimental evaluation compares our proposal to widely adopted algorithms for anomaly detection based on clustering and statistical approaches. We also compare the results to ROBPCA \cite{hubert2005robpca}, which is a method that also relies on robust estimates with adjusted outlyingness based on robust skewness.

The main contribution of this work is the proposal of a novel semi-supervised and unsupervised method for anomaly detection in skewed and imbalanced data, with results of experimental evaluation on simulated and real data sets.

This Chapter is organized as follows. In Section \ref{sec:4_relatedworks}, it is conducted a literature review of network anomaly detection, botnet detection, and imbalanced learning. Section \ref{sec:4_datamodel} presents the data model and the evaluated data set. Section \ref{sec:4_proposal} describes the proposed approach for network attack detection. Section \ref{sec:4_experiments} discusses the experimental validation, Section \ref{sec:4_simulated_result} presents the results of the simulated scenarios and the Section \ref{sec:4_ctu13_result} describes the results for anomaly detection on CTU-13 data set. Finally, Section \ref{sec:4_conclusion} draws the conclusions and the suggestions for future work.

@@

% network anomaly detection
Network Anomaly detection has emerging as an important approach for securing communication networks and to deal with the increasing number of network attacks. Bhuyan \emph{et al.} \cite{bhuyan2014network} provide an overview of facets of network anomaly detection, present attacks normally encountered by network intrusion detection systems, and categorize existing network anomaly detection methods and systems based on the underlying techniques. Ahmed \emph{et al.} \cite{ahmed2016survey} present an analysis of four major categories of anomaly detection techniques, which include classification, statistical, information theory and clustering. Moustafa \emph{et al.} \cite{moustafa2019holistic} discuss aspects of anomaly-based Network Intrusion Detection Systems (NIDSs), describing details of cyber-attacks and new solutions for anomaly detection, and provides a benchmark data sets for training and validating approaches for network anomaly detection.

@@

Acarali \emph{et al.} \cite{acarali2016survey} surveys network-based detection approaches for HTTP-based botnets, and discuss the traffic-based features used to detect bot traffic and presents an abstraction of the main types of features related to protocols and OSI layers. Wang \emph{et al.} \cite{Wang2018ddosbotnetssurvey} present an analysis based on 50,704 different Internet DDoS attacks originated of 674 botnets from 23 different botnet families with a total of 9,026 victim belonging to 1,074 organizations in 186 countries. Their analysis reveals that geolocation of the attacking sources follows patterns and enables source prediction, and highlights that multiple attacks to the same target also exhibit strong patterns of inter-attack time interval, also presents that there is a trend for different botnets to launch DDoS attacks targeting the same victim, simultaneously or in turn.

@@

Several approaches for network attack detection uses the KDD 99 \cite{ahmed2016survey,osanaiye2016distributed,bhuyan2014network} data sets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 data set are criticized by the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled data sets currently in use today by researchers \cite{osanaiye2016distributed,bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} data set is the refined version of the KDD 99 data set that removed the redundant data records, in order to avoid biased classifications. However, NSL-KDD data set maintain the limitations of the KDD 99 regarding volume and lacks on reproduction of recent network traffic and attacks.

The objective of Garcia \emph{et al.} \cite{garcia2014empirical} is to compare three botnet detection methods by means of a simple and reproducible methodology, by a good data set and by a new error metric. This paper evaluates some data sets for network anomaly detection, and surveys some approaches for botnet detection, and proposes two methods (BClus and CAMNEP) for botnet detection, comparing results to BotHunter \cite{gu2007bothunter}.

Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by attack, normal and background labeled data, in an imbalanced distribution like in a real networks. The authors recommend scenarios for training and testing in order to avoid the use of traffic from a botnet family for training and testing, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Considering the adoption of unsupervised or semi-supervised approaches for anomaly detection, adopting the training and testing approach proposed by Garcia \emph{et al.}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training.

Wang and Paschalidis \cite{wang2017botnet} proposed a botnet detection approach based on anomaly and community detection, aiming for detecting botnets and identifying bots before the botnet becomes active. The first stage detects anomalies by leveraging large deviations of an empirical distribution. The second stage detects the bots using ideas from social network community detection in a graph that captures correlations of interactions among nodes over time. This work is compared to the BotHunter \cite{gu2007bothunter} for the CTU-13 botnet data set \cite{garcia2014empirical}.

% Anomaly detection by subspace learning
Traditional PCA-based anomaly detection models are not suitable for anomaly interpretation, as they judge whether a data instance is an anomaly or not based on the length of its projection on the abnormal subspace spanned by the less significant principal components, and there is no direct mapping between PCA’s dimensionality-reduced subspace and the original feature space \cite{ringberg2007sensitivity}. However, to overcome the above mentioned limitations, some approaches based on PCA have been proposed for network anomaly detection. Callegari \emph{et al.} \cite{callegari2011novel} proposed a PCA-based method for identifying the network traffic flows responsible for an anomaly detected at the aggregate level, by means of a separation of normal and anomalies according to principal components (normal) and remaining (anomalies). Lee \emph{et al.} \cite{Lee2013} presented OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. Vieira \emph{et al.} \cite{vieira2017model} proposed a framework that applies Model Order Selection (MOS) for detection of time frames under attack and uses similarity analysis to extract details and detect the time and ports under attack.

% Anomaly detection by robust subspace learning
The problem of PCA or subspace learning for outlier corrupted data is called Robust Principal Component Analysis (RPCA) or robust subspace learning \cite{candes2011robust, vaswani2018robust}. RPCA aims to be resilient to outliers by means of a robust subspace learning \cite{vaswani2018robust} for outlier corrupted data, decomposing a given data matrix $\pmb{X}$ into the sum of a low rank matrix $\pmb{L}$, whose column subspace gives the principal components, and a sparse matrix $\pmb{S}$, which refers to outliers’ matrix. This definition is also referred to as the sparse + low rank ($\pmb{S + LR}$) formulation \cite{vaswani2018robust}. We refer to \cite{lerman2018overview} and \cite{vaswani2018robust} for more details regarding robust subspace learning.

RPCA is been mainly applied to computer vision, in problems of robust subspace tracking and robust subspace recovery. However, RPCA has also been adopted for general outlier detection \cite{hubert2005robpca,hubert2009robustskewed,cherapanamjeri2017thresholding,zhou2017anomaly,NetflixSurus} and for anomaly detection on network traffic \cite{pascoal2012robust}. ROBPCA \cite{hubert2005robpca} intends to identify outliers using PCA from robust estimates of mean and covariance matrix, to reduce the data dimensions and plotting the orthogonal distances versus the robust score distances, to flag an outlier map. However, ROBPCA flags many points as outlying when the original data is skewed. Therefore, Hubert \emph{et al.} \cite{hubert2009robustskewed} proposed ROBPCA-AO, which improves ROBPCA for problems with skewed data, by means of an adjusted outlyingness based on robust skewness. Hubert \emph{et al.} \cite{hubert2009robustskewed} evaluated ROBPCA-AO for real and simulated data sets, but it is not clear if this method was evaluated for very skewed and imbalanced data sets, such as in the network attack detection problem.

@@

Mahalanobis Distance (MD) is a generalized distance which is useful for determining the similarity between an unknown sample and a collection of known samples, by considering the covariance between the variables and their mean values. The MD is a measure of the distance between a vector $\pmb{x}$ and a distribution $\pmb{X}$, introduced by P. C. Mahalanobis in 1936 \cite{mahalanobis1936md}. It is a multi-dimensional generalization for measuring how many standard deviations away $\pmb{x}$ is from the mean $\bar{\pmb{x}}$ and covariance $\bar{\pmb{\Sigma}}$ of $\pmb{X}$. MD has been used for distance based anomaly detection with robust estimates in many areas, assuming that the Mahalanobis Distance between robust estimates and new observations can reveal anomalies.

We propose an approach based on the distances of moments computed from learned robust subspace and new observations, for anomaly detection on skewed and imbalanced data sets, and evaluate our approach for network attack detection. The proposed approach relies on learning robust subspace from supposed normal traffic for estimating the moments (mean, skewness and kurtosis). Thus, the anomaly detection for new observations should evaluate the Mahalanobis Distance between the robust moments and the moments of new observations, in a semi-supervised fashion and without the computational cost of new robust subspace learning for new observations, or in a unsupervised approach, where the robust subspace is learned from the contaminated data and the distance analysis is between the robust moments and the contaminated moments.

@@

In this paper, scalars are denoted by italic letters (\emph{a, b, A, B, $\alpha$, $\beta$}), vectors by lowercase bold letters ($\pmb{a}, \pmb{b}$), matrices by uppercase bold letters ($\pmb{A}$, $\pmb{B}$), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix $\pmb{A}$. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively. The Frobenius norm is denoted as $\left\| \mathord{\cdot} \right\|_F$, while $\left\| \mathord{\cdot} \right\|_*$ denotes the nuclear norm of a matrix and $\left\| \mathord{\cdot} \right\|_1$ means the sum of the absolute values of matrix entries. We also define the operator $\langle \mathord{\cdot} \rangle$, that is the standard trace inner product, and $[ \mathord{\cdot} ]^s$, which denotes the $n$ largest values of a vector.

This section also presents a description of the simulated data model in Subsection \ref{sec:4_simulation} and in Subsection \ref{sec:4_CTU-13} we describe the data model for the CTU-13 data set.

\subsection{The Simulated data set}
\label{sec:4_simulation}

In order to analyse the null hypotheses $H_{1rob.sub}$ and $H_{0rob.sub}$, which evaluate if the robust subspace learning can improves the anomaly detection from imbalanced and skewed data, we create two simulated data sets characterized by skewed and heavy tailed distributions, and create a simulated Gaussian distributed data set, in order to also analyse the detection rate for not skewed and not heavy tailed distributions.

We selected Pareto (with scale $a=3$, mode $m=1$ and $\mu=1$) and Lognormal (with mean $\mu=0$ and standard deviation $\sigma=1$) distributions, denoted respectively as $\pmb{Y}_p$, while $\pmb{Y}_l$, to simulate normal events and to evaluate the anomaly detection on skewed data performed by our proposals in comparison to widely adopted algorithms for outlier detection, considering that these distributions are skewed and heavy-tailed, and have been adopted to characterize network traffic of the Internet and data centers \cite{benson2010network,leon2017probability}. We also adopt the Gaussian distributed data set $\pmb{Y}_g$ to simulate normal events with mean $\mu=0$ and unitary standard deviation $\sigma=1$.

The addictive anomalies for Pareto and Lognormal distributions are white Gaussian noise with mean $\mu=0$ and unitary standard deviation $\sigma=1$. The Pareto distribution contaminated by Gaussian noise is denoted as $\pmb{Y}_p^c$, while $\pmb{Y}_l^c$ denotes the Lognormal distribution contaminated by Gaussian noise. The $\pmb{Y}_p^c$ and $\pmb{Y}_l^c$ are depicted by the Figures \ref{fig:4.01} and \ref{fig:4.02}, in order to provide a visual example one variable of $\pmb{Y}_p^c$ and $\pmb{Y}_l^c$.

\begin{figure}[!htb]
	\centering
	\subfigure[Pareto and Gaussian Anomalies]{%
		\includegraphics[width=0.48\textwidth]{figures/ch4/4_Xpg2.png}
		\label{fig:4.01}
	}
	\subfigure[Lognormal and Gaussian Anomalies]{%
		\includegraphics[width=0.48\textwidth]{figures/ch4/4_Xlogng2.png}
		\label{fig:4.02}
	}
	\caption[Examples of skewed and heavy tailed distributions]{Examples of skewed and heavy tailed distributions}
	\label{fig:4.03}
\end{figure}

An uniform noise between $-6$ and $6$ is used to add anomalies into the Gaussian distributed data set, and $\pmb{Y}_g^c$ denotes the contaminated Gaussian data contaminated by an uniform distribution. The figure \ref{fig:4.04} shows an example of the Gaussian distribution of normal observations and the uniform distribution to simulate the addition of anomalies.

\begin{figure}[h!]
	\centering
	\includegraphics[width=8cm]{figures/ch4/4_Xgu2.png}
	\caption{Example of Gaussian and Uniform Anomalies}
	\label{fig:4.04}
\end{figure}

Each contaminated data set is composed by a number of legitimate observations and contaminated samples. We evaluate contamination rates $c$ between 1\% and 50\%, to simulate the imbalanced data of anomaly detection problems during our experiments. This data sets simulate a total of 2400 events for each scenario with contaminated Pareto, Lognormal and Gaussian. Therefore, the number of legitimate observations is defined according to the contamination rate selected for each evaluation.

@@

\begin{itemize}
	\item \pmb{Attacks:} Click Fraud (CF), Port Scan (PS), Compiled and Controlled by Authors (US), SPAM and DDOS;
	\item \pmb{C\&C:} IRC, P2P and HTTP.
\end{itemize}

@@

We conducted and Exploratory Data Analysis (EDA) on the CTU-13 and observed that some features are skewed and present an overlapping between normal and anomalous flows, as can be seen in Figure \ref{fig:4.05} and Figure \ref{fig:4.06}, that present the distributions of TCP state of scenario 16 and the type of services from destination of scenario 10.

\begin{figure}[!htb]
	\centering
	\subfigure[State of scenario 16.]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/raw_distplot_capture20110816_State.png}
		\label{fig:4.05}
	}
	\subfigure[Destination Type of service of scenario 10.]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/raw_distplot_capture20110810_dTos.png}
		\label{fig:4.06}
	}
	\caption[Skewness and Overlapping]{Example of skewness and overlapping}
	\label{fig:4.07}
\end{figure}

However, it is not possible to observe a pattern on distributions of all the features and scenarios of CTU-13, as depicted by Figures \ref{fig:4.08} and \ref{fig:4.09}, that show the distributions of TCP states of the scenario 10 and source ports of scenario 16, and highlight the distributions of normal and anomalous flows.

\begin{figure}[!htb]
	\centering
	\subfigure[State of scenario 10.]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/raw_distplot_capture20110810_State.png}
		\label{fig:4.08}
	}
	\subfigure[Source Port of scenario 16.]{%
		\includegraphics[width=0.47\textwidth]{figures/ch4/raw_distplot_capture20110816_Sport.png}
		\label{fig:4.09}
	}
	\caption[Skewness and Overlapping]{Skewness and Overlapping}
	\label{fig:4.10}
\end{figure}

Due to the number of available features and the class overlapping between normal and anomalous flows, we performed an correlation analysis and an empirical cross validation in order to identify the best set of features for network attack detection. Therefore, we adopt the following features: state, destination type of service, destination port, source port, total number of packets, total number of bytes and number of bytes from the source.

\section{Moment Distances from Robust Subspace for Network Attack Detection}
\label{sec:4_proposal}

This section describes the proposed approach for network attack detection by means of a distance analysis between moments computed from a robust subspace and moments of new observations of network traffic.

Robust subspace learning can be defined as the decomposition of a given data matrix $\pmb{X}$ into the sum of a low rank matrix $\pmb{L}$, whose column subspace gives the principal components, and a sparse matrix $\pmb{S}$, with outliers or noise. We propose to learn a robust subspace $\pmb{L}$ from the normal traffic $\pmb{X}^s$ for computing $\pmb{L}^s$, $\pmb{S}^s$ and the robust moments, i.e. the mean $\bar{\pmb{x}}$, skewness $\bar{\pmb{s}}$ and kurtosis $\bar{\pmb{k}}$, in order to evaluate the distance $\pmb{d}$ between moments of new observations $\pmb{X}^c$ and moments computed from the learned robust subspace, in order to classify as anomalous the observations with the largest distances.
